---
title: "Individual-Level GxE Simulation — Reproducing Zhu et al. (2024)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Individual-Level GxE Simulation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE  # vignette requires MRGxE installation
)
```

## Overview

This vignette documents the new individual-level simulation functions in MRGxE (v0.3.0+). These functions reproduce the exact simulation methodology from:

> Zhu, X., Yang, Y., Lorincz-Comi, N. et al. *An approach to identify gene-environment interactions and reveal new biological insight in complex traits.* **Nature Communications** 15, 3385 (2024). doi:10.1038/s41467-024-47806-3

based on the original simulation code at [https://github.com/harryyiheyang/MR.GxE.Code](https://github.com/harryyiheyang/MR.GxE.Code) (Zenodo: 10.5281/zenodo.10815731).

## New Functions

| Function | Purpose | Paper Figure |
|----------|---------|-------------|
| `simulate_gxe_fig2ad_exact()` | 200-variant, no mediation, binary environment | Fig 2A-D |
| `simulate_gxe_fig2ef_exact()` | 20-variant, with/without mediation | Fig 2E-F, Supp. Fig 10 |
| `compute_gxe_effects_individual()` | Compute GWAS+GWIS effects from ind.-level data | Building block |
| `generate_gxe_phenotype()` | Generate phenotypes under 3 GxE models | Building block |
| `convert_to_mrgxe_format()` | Convert simulation output to pipeline input | Utility |
| `run_full_simulation_pipeline()` | End-to-end simulation + screening | Integration |
| `validate_simulation()` | Quick validation against paper expectations | Testing |
| `plot_fig2ad()` | Plot theta, beta3, type I error, power | Visualization |
| `plot_fig2ef_power()` | Plot power/type I error curves | Visualization |

## Design 1: Fig 2A-D (Many Variants, No Mediation)

This design generates M=200 independent variants with a binary environment variable. It tests T_MRGxE against T_Direct under varying environment distributions between the GWAS and GWIS cohorts.

### Core assumptions

- **No mediation**: E does not depend on G
- **Binary environment**: E ~ Binom(1, p_E), standardised to unit variance
- **Interaction only on variant 1**: beta_3 = 0.005
- **theta estimated from variants 3..M**: avoids correlation with the interaction variant
- **Three sign scenarios**: same direction (s=+1), opposite (s=-1), no main effect (s=0)

### Quick example

```{r}
library(MRGxE)

# Small-scale run for demonstration
set.seed(42)
sim_ad <- simulate_gxe_fig2ad_exact(
  n_variants = 200,
  n_gwas     = 160000,
  n_gwis     = 80000,
  mu_e       = c(0.1, 0.3, 0.5),
  ssign      = c(1, -1, 0),
  n_sim      = 50,           # use 100+ for publication
  verbose    = TRUE
)

# Check results
# Mean theta should be ~1
mean(sim_ad$theta, na.rm = TRUE)

# Plot theta estimates
plot_fig2ad(sim_ad, "theta")

# Plot power curves
plot_fig2ad(sim_ad, "power")
```

### Full-scale replication

To reproduce the paper's Fig 2A-D exactly:

```{r}
# WARNING: This takes several hours even with parallelisation
sim_ad_full <- simulate_gxe_fig2ad_exact(
  n_variants = 200,
  n_gwas     = 160000,
  n_gwis     = 80000,
  mu_e       = seq(0.05, 0.5, by = 0.05),  # 10 values
  ssign      = c(1, -1, 0),
  n_sim      = 300,         # paper used ~3000
  n_cores    = parallel::detectCores() / 2
)
```

## Design 2: Fig 2E-F (20 Variants, Mediation Scenarios)

This design tests 20 independent variants under six scenarios combining type I error estimation and power analysis, with and without mediation through the environment factor. Environment distribution differs between GWAS and GWIS cohorts (mean μ_E for GWIS, 2×μ_E for GWAS-only).

### Six scenarios

| # | Type | Mediation | E contrib. | γ (env scale) |
|---|------|-----------|-----------|---------------|
| 1 | Type I | None | 1% | 1 |
| 2 | Type I | 1% | 1% | 1 |
| 3 | Type I | 5% | 5% | √5 |
| 4 | Power | None | 1% | 1 |
| 5 | Power | 1% | 1% | 1 |
| 6 | Power | 5% | 5% | √5 |

### Quick example

```{r}
set.seed(123)
sim_ef <- simulate_gxe_fig2ef_exact(
  n_variants = 20,
  n_gwas     = c(20000, 100000, 200000, 300000),
  n_gwis     = 20000,
  n_sim      = 100,          # use 1000 for publication
  scenarios  = 1:6,
  verbose    = TRUE
)

# View results
sim_ef$results[, , "T_MRGxE"]

# Plot
plot_fig2ef_power(sim_ef, "type1")
plot_fig2ef_power(sim_ef, "power")
```

## Three Phenotype Models

The `generate_gxe_phenotype()` function implements the three phenotype models from the paper:

```{r}
n <- 20000
g <- rbinom(n, 2, 0.3) / sqrt(2 * 0.3 * 0.7)
e <- rnorm(n, mean = 1, sd = 1)

# Model 1: No mediation, no interaction
y1 <- generate_gxe_phenotype(cbind(g), e, model = "none")

# Model 2: Mediation only (E depends on G, but no GxE)
y2 <- generate_gxe_phenotype(cbind(g), e, model = "mediation_only")

# Model 3: Mediation + interaction
y3 <- generate_gxe_phenotype(cbind(g), e, model = "mediation_interaction")
```

## Integration with the MRGxE Pipeline

Once you have individual-level simulation results, you can feed them into the MRGxE screening pipeline:

```{r}
# Generate data
sim <- simulate_gxe_fig2ad_exact(
  n_variants = 200, n_sim = 1, verbose = FALSE
)

# Convert to MRGxE format
input_data <- convert_to_mrgxe_format(sim)

# Run screening pipeline
result <- run_pipeline(
  effect_a_path = NULL,  # use in-memory data
  effect_b_path = NULL,
  design_type = "gxe",
  gwas_data = input_data$gwas,
  gwis_data = input_data$gwis,
  rho = input_data$rho,
  verbose = TRUE
)
```

## Validation

Run a quick validation to verify that the simulation produces results consistent with the paper:

```{r}
validate_simulation(n_sim = 100)
```

This checks:
1. Mean theta estimate ≈ 1.0
2. Type I error ≈ 0.05 under the null
3. Power follows expected patterns (MRGxE generally outperforms Direct test)

## Key Differences from the Original `simulate_gxe_summary()`

The existing `simulate_gxe_summary()` and `create_example_data()` functions generate **summary statistics directly** — they create BETA, SE, and P values without ever generating individual-level data. These are useful for quick testing of the pipeline but do not reproduce the paper's simulation methodology.

The new functions in this vignette:

- **Generate individual-level data** matching the paper's equations exactly
- **Compute GWAS/GWIS effects from scratch** using the same method as the paper
- **Reproduce the paper's simulation parameters** (sample sizes, variant counts, environment distributions)
- **Include mediation modelling** (Fig 2E-F)
- **Support missing scenarios** (varied environment distributions between cohorts, sample overlap)

## References

1. Zhu, X. et al. (2024) Nature Communications, 15:3385. [doi:10.1038/s41467-024-47806-3](https://doi.org/10.1038/s41467-024-47806-3)
2. Original simulation code: [github.com/harryyiheyang/MR.GxE.Code](https://github.com/harryyiheyang/MR.GxE.Code)
3. MRGxE package: [github.com/Alerthater/-MRGxE](https://github.com/Alerthater/-MRGxE)
