---
title: "MRGxE: TMRGxE Gene-Environment Interaction Screening"
author: "MRGxE Authors"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{MRGxE: TMRGxE GxE Screening}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)
```

## Introduction

The MRGxE package implements the TMRGxE approach for detecting gene-environment
interactions (G×E) using GWAS and GWIS summary statistics, based on the
Mendelian randomization framework described in Zhu et al. (2024) *Nature
Communications*.

The core insight is that **GWAS marginal effects** (\(\hat{\alpha}\)) and **GWIS
main effects** (\(\hat{\beta}_1\)) have a linear relationship:

\[
\hat{\alpha} = \theta \hat{\beta}_1 + \text{deviation}
\]

Variants with G×E or mediation effects show a systematic deviation from this
regression line, analogous to horizontal pleiotropic variants in MR analysis.

## Installation

```{r installation, eval = FALSE}
# Install IMRP (required dependency)
remotes::install_github("XiaofengZhuCase/IMRP")

# Install MRGxE
install.packages("MRGxE", repos = NULL, type = "source")
```

## Quick Start with Simulated Data

```{r quickstart, results = 'hide'}
library(MRGxE)

# Step 0: Create simulated example data
sim <- create_example_data(seed = 42)

# Step 1: Harmonize and QC
harm <- harmonize_gwas_gwis(sim$gwis, sim$gwas)
```

```{r qc-summary}
# Check QC counts
harm$qc_counts
```

```{r iv-selection}
# Step 2: Select instrumental variables
iv <- select_instruments(harm$cleaned, p_col = "P")
```

```{r rho}
# Step 3: Estimate rho
rho <- estimate_rho(harm$cleaned)
```

```{r tmrgxe}
# Step 4 & 5: Run TMRGxE genome-wide screen
screen <- run_tmrgxe(harm$cleaned, iv$instruments, rho$rho)
```

## Using the Pipeline Wrapper

For a complete end-to-end analysis, use `run_pipeline()`:

```{r pipeline, eval = FALSE}
result <- run_pipeline(
  gwas_path = "my_gwas.txt",
  gwis_path = "my_gwis.txt",
  col_map = list(
    snp = "RSID", chr = "CHR", bp = "BP",
    a1 = "A1", a2 = "A2",
    gwas_beta = "BETA", gwas_se = "SE", gwas_n = "N",
    gwis_beta = "BETA_MAIN", gwis_se = "SE_MAIN", gwis_n = "N_GWIS"
  ),
  env_variable = "BMI",
  output_dir = "./GxE_results"
)
```

Or with a configuration file:

```{r pipeline-config, eval = FALSE}
result <- run_pipeline(config = "config.txt")
```

## Visualizing Results

```{r plots, warning = FALSE}
# Manhattan plot
plot_tmrgxe_manhattan(screen$results)

# QQ plot
plot_tmrgxe_qq(screen$results)

# Scatter plot
plot_tmrgxe_scatter(screen$results, imrp_result = screen$imrp)
```

## Generating Reports

```{r report, eval = FALSE}
# Generate markdown report
report <- generate_report(
  tmrgxe_result = screen,
  harmonize_result = harm,
  iv_result = iv,
  rho_result = rho
)
cat(report, sep = "\n")
```

## Module Reference

| Module | Function | Description |
|--------|----------|-------------|
| Data I/O | `import_summary_stats()` | Read and standardize GWAS/GWIS files |
| QC | `harmonize_gwas_gwis()` | Merge, align alleles, filter variants |
| IV Selection | `select_instruments()` | Select independent instruments |
| Rho | `estimate_rho()` | Estimate GWAS-GWIS correlation |
| MR | `run_imrp()` | Run IMRP estimation |
| Screen | `run_tmrgxe()` | Genome-wide TMRGxE test |
| Validation | `two_step_validation()` | Two-step G×E validation |
| Visualization | `plot_tmrgxe_*()` | Manhattan, QQ, scatter plots |
| Reporting | `generate_report()` | Markdown results report |
| Pipeline | `run_pipeline()` | Complete end-to-end pipeline |

## Migration Guide

To apply this method to a **new trait × environment combination**, you need:

1. **GWAS marginal effects** for the trait (large sample preferred)
2. **GWIS main effects** from a model that includes the environment E as a
   covariate (\(Y = \beta_0 + \beta_1 G + \beta_2 E + \epsilon\))

The environment E can be:
- **Binary**: smoking status, alcohol consumption
- **Multi-category**: education level (≥3 categories), encoded as multiple
  dummy variables
- **Continuous**: BMI, blood pressure, age

### Adjustable Parameters

- `iv_p_threshold`: Relax if GWAS is underpowered (e.g., 1e-6)
- `tmrgxe_threshold`: Genome-wide significance (typically 5e-8)
- `qc_max_eaf_diff`: Stricter (0.1) or looser (0.2) EAF concordance
- `ld_method`: Use "plink" for real data with LD reference, "distance" as proxy
- `min_instruments`: Relax if few genome-wide significant loci exist
